---
title: "Milestone 1 – Member 3: Data Understanding (EDA & Quality)"
author: "Team HDPSA"
output:
  html_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview
This document loads all provided datasets, reports dimensions, variable types, missing values, duplicate rows, outliers (basic), and summary statistics. It saves CSV summaries into `outputs/`.

```{r libraries}
library(readr)
library(dplyr)
library(purrr)
library(stringr)
```

```{r paths}
input_dir <- "../Project Datasets"
output_dir <- "../outputs"
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
```

## Load datasets
```{r load}
csv_files <- list.files(input_dir, pattern = "\\.csv$", full.names = TRUE)
stopifnot(length(csv_files) > 0)

# Named list of tibbles
name_from_path <- function(p) gsub("\\.csv$", "", basename(p))
raw_datasets <- csv_files |> setNames(nm = vapply(csv_files, name_from_path, character(1))) |> 
  lapply(readr::read_csv, show_col_types = FALSE)
```

## Structure and dimensions
```{r structure}
structure_tbl <- purrr::imap_dfr(raw_datasets, function(df, nm){
  tibble::tibble(
    dataset = nm,
    n_rows = nrow(df),
    n_cols = ncol(df)
  )
})
readr::write_csv(structure_tbl, file.path(output_dir, "structure_summary.csv"))
structure_tbl
```

## Variable types
```{r types}
coltypes_tbl <- purrr::imap_dfr(raw_datasets, function(df, nm){
  tibble::tibble(
    dataset = nm,
    column = names(df),
    type = vapply(df, function(x) class(x)[1], character(1))
  )
})
readr::write_csv(coltypes_tbl, file.path(output_dir, "column_types.csv"))
coltypes_tbl |> dplyr::count(type, sort = TRUE)
```

## Missing values and duplicates
```{r missing-duplicates}
missing_tbl <- purrr::imap_dfr(raw_datasets, function(df, nm){
  tibble::tibble(
    dataset = nm,
    total_cells = nrow(df) * ncol(df),
    total_missing = sum(is.na(df)),
    pct_missing = round(100 * total_missing / pmax(1, total_cells), 2)
  )
})
readr::write_csv(missing_tbl, file.path(output_dir, "missingness_summary.csv"))
missing_tbl

duplicates_tbl <- purrr::imap_dfr(raw_datasets, function(df, nm){
  dup_count <- sum(duplicated(df))
  tibble::tibble(dataset = nm, duplicate_rows = dup_count)
})
readr::write_csv(duplicates_tbl, file.path(output_dir, "duplicates_summary.csv"))
duplicates_tbl
```

## Summary statistics (numerical columns)
```{r numeric-summaries}
num_summary_tbl <- purrr::imap_dfr(raw_datasets, function(df, nm){
  num_df <- dplyr::select(df, where(is.numeric))
  if(ncol(num_df) == 0) return(tibble::tibble())
  num_df |>
    summarize(across(everything(), list(
      n = ~sum(!is.na(.x)),
      mean = ~mean(.x, na.rm = TRUE),
      sd = ~sd(.x, na.rm = TRUE),
      min = ~min(.x, na.rm = TRUE),
      q25 = ~quantile(.x, 0.25, na.rm = TRUE),
      median = ~median(.x, na.rm = TRUE),
      q75 = ~quantile(.x, 0.75, na.rm = TRUE),
      max = ~max(.x, na.rm = TRUE)
    ), .names = "{.col}.{.fn}")) |>
    mutate(dataset = nm, .before = 1)
})
readr::write_csv(num_summary_tbl, file.path(output_dir, "numeric_summary.csv"))
num_summary_tbl |> dplyr::slice_head(n = 10)
```

## Basic outlier detection (IQR rule, counts per numeric column)
```{r outliers}
outliers_tbl <- purrr::imap_dfr(raw_datasets, function(df, nm){
  num_df <- dplyr::select(df, where(is.numeric))
  if(ncol(num_df) == 0) return(tibble::tibble())
  purrr::imap_dfr(num_df, function(col, cname){
    q1 <- quantile(col, 0.25, na.rm = TRUE)
    q3 <- quantile(col, 0.75, na.rm = TRUE)
    iqr <- q3 - q1
    lower <- q1 - 1.5*iqr
    upper <- q3 + 1.5*iqr
    tibble::tibble(dataset = nm, column = cname,
                   outlier_count = sum(col < lower | col > upper, na.rm = TRUE))
  })
})
readr::write_csv(outliers_tbl, file.path(output_dir, "outliers_summary.csv"))
outliers_tbl |> dplyr::arrange(desc(outlier_count)) |> dplyr::slice_head(n = 10)
```

## Notes
- Use these summaries to inform cleaning and transformations in Milestone 2.
- Member 4 can build visuals from `outputs/` or reuse this document to render charts.
